Data Descriptive: Corresponding author journal publishing activities 2014 - 2018

### 2019-09-13 analysis/004_kb_fetch_publisher.Rmd
adding file name as suggested by  @gitti1

```{r}
rp_df_oa %>%
  select(
    issn_wos = ISSN,
    publication_year = PUBYEAR,
    articles = PUBS,
    journal_title, publisher, oa_journal,
    issn_l = `ISSN-L`
    ) %>%
  write_csv("../data/publisher_14_18.csv")
```

@@

```{r}
readr::read_csv("../data/publisher_14_18.csv") %>%
  writexl::write_xlsx("../data/publisher_14_18.xlsx")
```

>>>>

```{r}
rp_df_oa %>%
  select(
    issn_wos = ISSN,
    publication_year = PUBYEAR,
    articles = PUBS,
    journal_title, publisher, oa_journal,
    issn_l = `ISSN-L`
    ) %>%
  write_csv("../data/journal_publisher_14_18.csv")
```

@@

```{r}
readr::read_csv("../data/journal_publisher_14_18.csv") %>%
  writexl::write_xlsx("../data/journal_publisher_14_18.xlsx")
```

### 2019-09-13 analysis/paper.Rmd
adding file name as suggested by  @gitti1

Only a few openly available datasets focusing on the scholarly publishing industry exist. For example, @haustein_2015 shared an aggregated dataset that contains the market share of publishers in terms of articles indexed in the Web of Science between 1973-2013. Using this data, the authors were able to reveal a growing market concentration among a few publishing houses [@Larivi_re_2015]. @Palzenberger_2015 provided a dataset presenting the number and proportion of scholarly articles per country of affiliation including a breakdown per author role for the period 2004-2014. Published in 2015, this dataset, which was obtained from the Web of Science in-house database from the German Competence Center for Bibliometrics, formed the empirical basis for demonstrating that a large-scale transition from a subscription-based to open access business model is feasible based on the current total spending on scholarly journals [@Schimmer_2015]. Since then, analyzing the productivity of corresponding authors in terms of articles published in dedicated journal portfolios has been perceived as a critical component of open access negotiations.

There have also been analytical applications for such datasets. Major bibliometric databases including the Web of Science, Scopus and Dimension have started to integrate open access status information. In these databases, Unpaywall data is often used to identify openly available full-texts at the article-level, while the Directory of Open Access Journals provides information about whether the journal is operating under a fully open access or toll-access business model. In its most recent edition, the Leiden Ranking presents open access indicators at the level of universities [@leiden]. In addition to university-specific data, the German Open Access Monitor, provides breakdowns per publisher and journal, as well as the inclusion of publications from non-university research institutions based in Germany [@oa_monitor].

@@

The data cannot only serve as a critical input for negotiations in the context of transformative agreements, but also in the quantitative study of the academic publishing industry and the open access uptake across countries. For instance, national transformative agreements with large publishers will likely have an impact on the open access share of a country in the near future. To demonstrate the dataset's potential significance, we will present two new findings using the dataset: a global map of publishers's toll-access market share, and an analysis of country of affiliation by publisher. Source code used for compiling the dataset is shared along with the resulting datasets.

@@

- Web of Science collections SCI, SSCI and AHCI
- Article types Original Articles and Reviews
- Country of affiliations
- ISSN
- Publication years 2014 - 2018 
- Author Roles

Our database retrieval reflected that it has become challenging to count corresponding author affiliations straight where one country receives full credit for the publication [@Huang_2011], because the Web of Science has begun to syetmatically keep track of more than one corresponding author recently (see Figure 1). We also found that the Web of Science did record more than one corresponding author for 6.2% of indexed articles published between 2014 - 2018. Moreover, a  5% of corresponding authors author was internationally co-located (see Figure 2). These results support some concerns about how to allocate open access spending based on corresponding authorships [@Gumpenberger_2018]. Here, we decided to use whole counting where every original article or review was counted once per country of affiliation of the corresponding authors.

@@

After obtaining aggregated article counts at the country and journal level from the Web of Science, Crossref’s REST API was queried to retrieve publisher and journal titles. To control for developments of the publishing market resulting in name changes of publishers or journal titles over time, only the most frequent field name was used from the API facet counts. Drawing on ISSN-L, a jouranl identifier that links between different media versions of a journal, ISSN variants were obtained to improve the retrieval. Next, the open access status per journal was added to the dataset. This information was retrieved from a comprehensive list of open access journals indexed in the Web of Science, which joins open access journal information from various sources including the DOAJ, PubMed Central and the Directory of Open Access Scholarly Resources (ROAD) [@issn]. Again, ISSN-L were used for the matching procedure.

@@

Overall, our dataset consists of two separate tables, which are openly available at <https://github.com/subugoe/oa2020cadata> as comma-separated values (csv) files and Excel spreadsheets.

@@

```{r}
publisher_df <- readr::read_csv("../data/publisher_14_18.csv")
```

The first file, named `publisher_14_18`, contains information about the publishing market landscape for the period 2014-2018 in terms of the aggregated number of articles and reviews  by year and journal represented by the ISSN identifier indexed in the Web of Science. Table 1 describes the data variables in detail. We were able to retrieve publisher and journal titles from Crossref for `r publisher_df %>% filter(!is.na(publisher)) %>% distinct(issn_l) %>% nrow() / publisher_df %>% distinct(issn_l) %>% nrow() * 100` %
 of all journals, representing `r publisher_df %>% filter(!is.na(publisher)) %>% .$articles %>% sum() / sum(publisher_df$articles) * 100` % of the overall article volume. `r publisher_df %>% filter(oa_journal == TRUE) %>% distinct(issn_l) %>% nrow()` out of `r publisher_df %>% filter(!is.na(publisher)) %>% distinct(issn_l) %>% nrow()`
 investigated journals with `ISSN-L` were identified as fully open access journals. In total, the dataset provides information for `r sum(publisher_df$articles)` records indexed in the Web of Science with journal information in the five-years period 2014-2018.

Data Schema `publisher_14_18`:

@@

Based on the dataset `publisher_14_18` the global market share of publishing houses in terms of articles and reviews indexed in the Web of Science can be analyzed for the period 2014-2018. Table 1 shows the overall market share for the ten largest publishers. Figure 1 presents a breakdown by journal business model, differentiating between toll access and fully open access.

```{r, results='asis'}
my_df <- readr::read_csv("../data/publisher_14_18.csv")
tt_all <- my_df %>% 
  group_by(publisher) %>%
  summarise(n = sum(articles)) %>%
  mutate(prop = n /sum(n))  %>%
  mutate(publisher = ifelse(is.na(publisher), "Other", publisher)) %>%
  arrange(desc(n)) %>%
  mutate(publisher_group = fct_other(publisher, keep = publisher[1:11])) %>%
  ungroup() %>%
  group_by(publisher_group) %>%
  summarise(n = sum(n), prop = sum(prop)) %>%
  mutate(publisher_group = fct_reorder(publisher_group, n,)) %>%
  mutate(publisher_group = fct_relevel(publisher_group, "Other")) %>%
  arrange(desc(publisher_group))
tt_all %>%
  mutate(prop = prop * 100) %>%
  knitr::kable(caption = "Publisher Market Share 2014 - 2018 in terms of articles and reviews indexed in the Web of Science", col.names = c("Publisher", "Articles and Reviews published", "Percentage"))
```

>>>>

Only a few openly available datasets focusing on the scholarly publishing industry exist. For example, @haustein_2015 shared an aggregated dataset that contains the market share of publishers in terms of articles indexed in the Web of Science between 1973-2013. Using this data, the authors were able to reveal a growing market concentration among a few publishing houses [@Larivi_re_2015]. @Palzenberger_2015 provided a dataset presenting the number and proportion of scholarly articles by country of affiliation including a breakdown by author role for the period 2004-2014. Published in 2015, this dataset, which was obtained from the Web of Science in-house database from the German Competence Center for Bibliometrics, formed the empirical basis for demonstrating that a large-scale transition from a subscription-based to an open access business model is feasible based on the current total spending on scholarly journals [@Schimmer_2015]. Since then, analyzing the productivity of corresponding authors in terms of articles published in dedicated journal portfolios has been perceived as a critical component of open access negotiations.

There have also been analytical applications for such datasets. Major bibliometric databases including the Web of Science, Scopus and Dimension have started to integrate open access status information. In these databases, Unpaywall data is often used to identify openly available full-texts at the article-level, while the Directory of Open Access Journals provides information about whether the journal is operating under a fully open access or toll-access business model. In its most recent edition, the Leiden Ranking presents open access indicators at the level of universities [@leiden]. In addition to university-specific data, the German Open Access Monitor, provides breakdowns by publisher and journal, as well as the inclusion of publications from non-university research institutions based in Germany [@oa_monitor].

@@

The data cannot only serve as a critical input for negotiations in the context of transformative agreements, but also in the quantitative study of the academic publishing industry and the open access uptake across countries. For instance, national transformative agreements with large publishers will increase opportunities to publish open access. They  will therefore likely have an impact on the open access share of a country with such agreements. To demonstrate the dataset's potential significance, we will present two new findings using the dataset: a global map of publishers's toll-access market share, and an analysis of country of affiliation by publisher. Source code used for compiling the dataset is shared along with the resulting datasets.

@@

- Web of Science collections SCI, SSCI and AHCI
- Article types Original Articles and Reviews
- Country of affiliations
- ISSN
- Publication years 2014 - 2018 
- Author roles, interpreting the role "reprint author" as corresponding authorship.

There has been some concerns to allocate open access spending based on corresponding authorships, because it has become challenging to count corresponding author affiliations straight, i.e. a counting method where one institution or country receives full credit for a publication [@Huang_2011, @Gumpenberger_2018]. Indeed, the Web of Science has begun to systematically keep track of more than one corresponding author recently (see Figure 1). We also found that the Web of Science did record more than one corresponding author for 6.2% of indexed articles published between 2014 - 2018. Moreover, a 5% of corresponding authors author was internationally co-located (see Figure 2). We therefore decided to use whole counting where every original article or review was counted once per country of affiliation of the corresponding author(s).

@@

After obtaining aggregated article counts at the level of countries and journals from the Web of Science, Crossref’s REST API was queried to retrieve publisher and journal titles. To control for developments of the publishing market resulting in name changes of publishers or journal titles over time, only the most frequent field name was used from the API facet counts. Drawing on ISSN-L, a journal identifier that links between different media versions of a journal, ISSN variants were obtained to improve the retrieval. Next, the open access status per journal was identified and added to the dataset. This information was retrieved from a comprehensive list of open access journals indexed in the Web of Science, which joins open access journal information from various sources including the DOAJ, PubMed Central and the Directory of Open Access Scholarly Resources (ROAD) [@issn]. Again, ISSN-L were used for the matching procedure.

@@

Overall, our dataset consists of two separate tables, which are openly available at (Link to Zenodo) as comma-separated values (csv) files and Excel spreadsheets.

@@

```{r}
publisher_df <- readr::read_csv("../data/journal_publisher_14_18.csv")
```

The first file, named `journal_publisher_14_18`, contains information about the publishing market landscape for the period 2014-2018 in terms of the aggregated number of articles and reviews  by year and journal represented by the ISSN identifier indexed in the Web of Science. Table 1 describes the data variables in detail. We were able to retrieve publisher and journal titles from Crossref for `r publisher_df %>% filter(!is.na(publisher)) %>% distinct(issn_l) %>% nrow() / publisher_df %>% distinct(issn_l) %>% nrow() * 100` %
 of all journals, representing `r publisher_df %>% filter(!is.na(publisher)) %>% .$articles %>% sum() / sum(publisher_df$articles) * 100` % of the overall article volume. `r publisher_df %>% filter(oa_journal == TRUE) %>% distinct(issn_l) %>% nrow()` out of `r publisher_df %>% filter(!is.na(publisher)) %>% distinct(issn_l) %>% nrow()`
 investigated journals with `ISSN-L` were identified as fully open access journals. In total, the dataset provides information for `r sum(publisher_df$articles)` records indexed in the Web of Science with journal information in the five-years period 2014-2018.

Data Schema `journal_publisher_14_18`:

@@

Based on the dataset `journal_publisher_14_18` the global market share of publishing houses in terms of articles and reviews indexed in the Web of Science can be analyzed for the period 2014-2018. Table 1 shows the overall market share for the ten largest publishers. Figure 1 presents a breakdown by journal business model, differentiating between toll access and fully open access.

```{r, results='asis'}
my_df <- readr::read_csv("../data/journal_publisher_14_18.csv")
tt_all <- my_df %>% 
  group_by(publisher) %>%
  summarise(n = sum(articles)) %>%
  mutate(prop = n /sum(n))  %>%
  mutate(publisher = ifelse(is.na(publisher), "Other", publisher)) %>%
  arrange(desc(n)) %>%
  mutate(publisher_group = fct_other(publisher, keep = publisher[1:11])) %>%
  ungroup() %>%
  group_by(publisher_group) %>%
  summarise(n = sum(n), prop = sum(prop)) %>%
  mutate(publisher_group = fct_reorder(publisher_group, n,)) %>%
  mutate(publisher_group = fct_relevel(publisher_group, "Other")) %>%
  arrange(desc(publisher_group))
tt_all %>%
  mutate(prop = prop * 100) %>%
  knitr::kable(caption = "Publisher Market Share 2014 - 2018 in terms of articles and reviews indexed in the Web of Science", col.names = c("Publisher", "Articles and Reviews published", "Percentage"))
```
